---
title: "Presentation"
author: "W. Evan Johnson, Ph.D., Ian Mc Farlane, Bram Stults"
date: "`r Sys.Date()`"
output: slidy_presentation
header-includes:
   - \usepackage{amsmath}
editor_options: 
  chunk_output_type: console
tables: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(readr)
library(Rsubread)
library(Rsamtools)
library(SummarizedExperiment)
library(edgeR)
library(DESeq2)
library(ComplexHeatmap)
library(TBSignatureProfiler)
library(umap)
library(tidyverse)
library(DT)
library(GSVA)
library(enrichR)
library(biomaRt)
library(rprojroot)
library(fs)
library(dplyr)
library(patchwork)
library(gridExtra)
library(ComplexHeatmap)

# define pathway libraries
dbs <- c("WikiPathways_2016","KEGG_2016","Panther_2016","Reactome_2016")

```

```{r read_in_blood, include=FALSE}
# read RNA expression counts and patient data
load(file="data/readMatrix_PBMC_BulkRNAseq_CVID.Rdata")

mt.counts.df <- read.csv("data/translated_genes.txt", row.names = 1)
groups <- sample.info$disease 

subs = c(UCVID = "Deficient",
         CVID = "Deficient",
         healthy = "Healthy",
         NCVID = "remove")


groups <- unname(subs[match(groups, names(subs))])

# store index of NCVID subjects
remove_NCVID_index <- which(groups == "remove")

# update group label vector
groups <- groups[groups != "remove"] 
groups <- factor(groups)


#updating dataframes
mt.counts.df <- mt.counts.df[,-remove_NCVID_index]
sample.info <- sample.info[-remove_NCVID_index,]
sample.info$Condition <- groups

#  re-store results
counts <- mt.counts.df
meta_data <- sample.info

# isolate group indicator variable
group <- meta_data$disease

# aggregate patient and expression data into a single object
se_cvid <- SummarizedExperiment(assays=list(counts=counts),
                     colData = meta_data)


# isolating treatment group experiments
treatments <- meta_data$treatment
ctrl_index <- which(treatments == "ctrl")
LPS_index <- which(treatments == "LPS")

counts_ctrl <- counts[,-LPS_index]
counts_LPS <- counts[,-ctrl_index]

meta_data_ctrl <- meta_data[-LPS_index,]
meta_data_LPS <- meta_data[-ctrl_index,]

se_cvid_ctrl <- SummarizedExperiment(assays=list(counts=counts_ctrl),
                     colData = meta_data_ctrl)
se_cvid_LPS <- SummarizedExperiment(assays=list(counts=counts_LPS),
                     colData = meta_data_LPS)

# create counts per million, log counts and log counts per million features
# and save them into new assays
se_cvid <- mkAssay(se_cvid, log = TRUE, counts_to_CPM = TRUE)
# same for control only
se_cvid_ctrl <- mkAssay(se_cvid_ctrl, log = TRUE, counts_to_CPM = TRUE)
# same for LPS treatment group only
se_cvid_LPS <- mkAssay(se_cvid_LPS, log = TRUE, counts_to_CPM = TRUE)
```

```{r read_in_nasal, include=FALSE}

# read RNA expression counts
counts <- read.table("data/features_combined.txt", sep="\t", 
                     header=T, row.names=1)
# read patient data
meta_data <- read.table("data/meta_data.txt", sep="\t", 
                     header=T, row.names=1)

# isolate group indicator variable
group <- meta_data$Condition
group <- ifelse(group == "Control", "Healthy", "Deficient")
meta_data$Condition <- group

# aggregate patient and expression data into a single object
se_immunedef <- SummarizedExperiment(assays=list(counts=counts),
                     colData = meta_data)

# create counts per million, log counts and log counts per million features 
# and save them into new assays.
se_immunedef <- mkAssay(se_immunedef, log = TRUE, counts_to_CPM = TRUE)

# display all assays
assays(se_immunedef)
```


# Nasal VS Control

## Visualization and Dimension reduction

```{r nasal_umap, include=FALSE}
# fit UMAP model to log cpm data
# note: transpose data to make it tidy
set.seed(1)
umap.defaults$n_neighbors=5
umap_out <- umap(t(assay(se_immunedef,"log_counts_cpm")), config = umap.defaults)

# define dataframe with UMAP dimensions and IDC
umap_plot <- as.data.frame(umap_out$layout) # X, Y, ...
umap_plot$Condition <- as.factor(se_immunedef$Condition) # color / group

# plot UMAP1 and UMAP2 by IDC
g1 <- umap_plot %>% ggplot(aes(x=V1, y=V2, color=Condition)) +
  geom_point(size=1.5) + xlab("UMAP1") + ylab("UMAP2") +
  geom_text(label=colnames(se_immunedef), nudge_y = 0.1,) +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("Nasal Samples")
```

```{r ctrl_umap, include=FALSE}
# fit UMAP model to log cpm data
# note: transpose data to make it tidy
set.seed(1)
umap.defaults$n_neighbors=5
umap_out <- umap(t(assay(se_cvid_ctrl,"log_counts_cpm")), config=umap.defaults)

# define dataframe with UMAP dimensions and IDC
umap_plot <- as.data.frame(umap_out$layout) # X, Y, ...
umap_plot$Condition <- as.factor(se_cvid_ctrl$Condition) # color / group

# plot UMAP1 and UMAP2 by IDC
g2 <- umap_plot %>% ggplot(aes(x=V1, y=V2, color=Condition)) +
  geom_point(size=1.5) + xlab("UMAP1") + ylab("UMAP2") +
  geom_text(label=colnames(se_cvid_ctrl), nudge_y = 0.1,) +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("PBMC - Control Samples")
```

```{r UMAPS, echo=FALSE}
combined_plot <- g1 + g2 + plot_layout(guides = "collect") & theme(legend.position = "bottom")

plot(combined_plot)
```


## DEGs (Deseq) Heat Maps

```{r nasal_deseq, include=FALSE}
# format data into a DESeq2-friendly Data Structure
counts <- assay(se_immunedef, "counts")
meta_data <- colData(se_immunedef)
Condition <- meta_data$Condition

dds <- DESeqDataSetFromMatrix(countData = counts, colData=meta_data, design=~Condition)
#colData is a data frame of demographic/phenotypic data

# filter for genes with significant expression  
# expression should be > 0, but > 100 should be better
dds<-dds[rowSums(counts(dds))>1,] #Gene Filtering

# fit negative binomial regression
dds<-DESeq(dds) #Performs estimation of size factors,dispersion, and negative binomial GLM f#itting

# extracting results, ordering them by adjusted p-value
res <- results(dds)[order(results(dds)[,6]),]

# store the name of the top 1000 most significant genes, per Dr. Johnson
nasal_deseq <- data.frame(res)
nasal_deseq1000 <- rownames(res)[1:1000]
```

```{r nasal_deseq_HA, include=FALSE}
# Make a Heatmap of DEGs

# All together: extract the log cpm of the top 1000 genes and store that in a matrix
# # get the name of the genes ordered by adjusted p-val
top_genes <- rownames(results(dds)[order(results(dds)$padj),])[1:250]
# # convert log cpm to a matrix structure sorted by the top genes
mat <- as.matrix(assay(se_immunedef, "log_counts_cpm"))[top_genes,] 
# Using first 500 genes to simplify

# normalize data row-wise
mat <- t(scale(t(mat)))

# store IDC into a single-column dataframe
df <- data.frame(Condition=colData(se_immunedef)$Condition) 

# create HeatMap with some annotations
ha <- HeatmapAnnotation(df=df, 
                        col=list(Condition=c("Deficient"="Red","Healthy"="Blue")),
                        annotation_name_side = "left")
HM1 <- Heatmap(mat,
               name = "Log CPM",
               show_row_names = F,
               show_column_names = F,
               row_dend_side = "left",
               top_annotation = ha,
               column_title = "Nasal Samples",
               heatmap_legend_param = list(direction = "horizontal", 
                                           title_gp = gpar(fontsize = 8), 
                                           labels_gp = gpar(fontsize = 8)))

```

```{r ctrl_deseq, include=FALSE}
# format data into a DESeq2-friendly Data Structure
counts <- assay(se_cvid_ctrl, "counts")
meta_data <- colData(se_cvid_ctrl)
Condition <- meta_data$Condition

dds <- DESeqDataSetFromMatrix(countData=counts, colData=meta_data, design=~Condition)
#colData is a data frame of demographic/phenotypic data

# filter for genes with significant expression  
# expression should be > 0, but > 100 should be better
dds<-dds[rowSums(counts(dds))>1,] #Gene Filtering

# fit negative binomial regression
dds<-DESeq(dds) #Performs estimation of size factors, dispersion, and negative binomial GLM f#itting

# extracting results, ordering them by adjusted p-value
res <- results(dds)[order(results(dds)[,6]),]
#res[1:10,]

# store the name of the top 1000 most significant genes, per Dr. Johnson
ctrl_deseq <- data.frame(res)
ctrl_deseq1000 <- rownames(res)[1:1000]
```


```{r ctrl_deseq_HA, include=FALSE}
# Make a Heatmap of DEGs

# All together: extract the log cpm of the top 1000 genes and store that in a matrix
# # get the name of the genes ordered by adjusted p-val
top_genes <- rownames(results(dds)[order(results(dds)$padj),])[1:250]
# # convert log cpm to a matrix structure sorted by the top genes
mat <- as.matrix(assay(se_cvid_ctrl,"log_counts_cpm"))[top_genes,] 

# normalize data row-wise
mat <- t(scale(t(mat)))

# store IDC into a single-column dataframe
df <- data.frame(Condition=colData(se_cvid_ctrl)$Condition) 

# create HeatMap with some annotations
ha <- HeatmapAnnotation(df=df, col=list(Condition=c("Healthy"="Blue","Deficient"="Red")))
HM2 <- Heatmap(mat,
               name = "Log CPM",
               show_row_names=F,
               show_column_names=F,
               row_dend_side = "right",
               top_annotation=ha,
               column_title = "PBMC - Control Samples",
               heatmap_legend_param = list(direction = "horizontal", 
                                           title_gp = gpar(fontsize = 8), 
                                           labels_gp = gpar(fontsize = 8)))

```

```{r deseq_Heatmaps}
h1 <- grid.grabExpr(draw(HM1,
                         heatmap_legend_side = "bottom",
                         annotation_legend_side = "bottom"))

h2 <- grid.grabExpr(draw(HM2,
                         heatmap_legend_side = "bottom",
                         annotation_legend_side = "bottom"))

# Arrange them separately
grid.arrange(h1, h2, ncol = 2)
```

## Combined DEGs (DeSeq)
```{r combined_results}
common_genes <- intersect(row.names(nasal_deseq), row.names(ctrl_deseq))

combined_results <- data.frame(gene = common_genes) %>%
    mutate(
        nasal_padj = nasal_deseq[common_genes,]$padj,
        ctrl_padj = ctrl_deseq[common_genes,]$padj,
        combined_chisq = -2*(log(nasal_padj) + log(ctrl_padj)),
        nasal_effect = nasal_deseq[common_genes,]$log2FoldChange,
        ctrl_effect = ctrl_deseq[common_genes,]$log2FoldChange,
    ) %>% 
    transmute(
        gene = gene,
        combined_p = 1 - pchisq(combined_chisq, 4),
        nasal_effect = ifelse(nasal_effect > 0, "Upregulated", "Downregulated"),
        ctrl_effect = ifelse(ctrl_effect > 0, "Upregulated", "Downregulated"),
        reg_agreement = ifelse(ctrl_effect == nasal_effect, "Same", "Opposite"),
    ) %>% 
    mutate(across(where(is.character), as.factor))%>% 
    filter(!is.na(combined_p)) %>%
    arrange(combined_p) %>% 
    slice_head(n = 250)

row.names(combined_results) <- combined_results$gene
combined_results$gene <- NULL

datatable(combined_results)
```

## Pathway Analysis - Any {.tabset}
```{r include = FALSE}
# based on the results of DEseq, cross-reference the gene with pathway databases to infer which pathways are the most significant

genes <- row.names(combined_results)
enriched <- enrichr(genes, dbs)
```

### DESeq2 WikiPathways

```{r}
datatable(enriched$WikiPathways_2016)
```

### DESeq2 KEGG

```{r}
datatable(enriched$KEGG_2016)
```

### DESeq2 Panther

```{r}
datatable(enriched$Panther_2016)
```

### DESeq2 Reactome

```{r}
datatable(enriched$Reactome_2016)
```

## Pathway Analysis - Same {.tabset}
```{r include=FALSE}
# based on the results of DEseq, cross-reference the gene with pathway databases to infer which pathways are the most significant

genes <- row.names(filter(combined_results, reg_agreement == "Same"))
enriched <- enrichr(genes, dbs)
```

### DESeq2 WikiPathways

```{r}
datatable(enriched$WikiPathways_2016)
```

### DESeq2 KEGG

```{r}
datatable(enriched$KEGG_2016)
```

### DESeq2 Panther

```{r}
datatable(enriched$Panther_2016)
```

### DESeq2 Reactome

```{r}
datatable(enriched$Reactome_2016)
```

## Pathway Analysis - Opposite {.tabset}
```{r include=FALSE}
# based on the results of DEseq, cross-reference the gene with pathway databases to infer which pathways are the most significant

genes <- row.names(filter(combined_results, reg_agreement == "Opposite"))
enriched <- enrichr(genes, dbs)
```
### DESeq2 WikiPathways

```{r}
datatable(enriched$WikiPathways_2016)
```

### DESeq2 KEGG

```{r}
datatable(enriched$KEGG_2016)
```

### DESeq2 Panther

```{r}
datatable(enriched$Panther_2016)
```

### DESeq2 Reactome

```{r}
datatable(enriched$Reactome_2016)
```


# Nasal VS LPS

## Visualization and Dimension reduction

```{r lps_umap, include=FALSE}
# fit UMAP model to log cpm data
# note: transpose data to make it tidy
set.seed(1)
umap.defaults$n_neighbors=5
umap_out <- umap(t(assay(se_cvid_LPS,"log_counts_cpm")), config=umap.defaults)

# define dataframe with UMAP dimensions and IDC
umap_plot <- as.data.frame(umap_out$layout) # X, Y, ...
umap_plot$Condition <- as.factor(se_cvid_ctrl$Condition) # color / group

# plot UMAP1 and UMAP2 by IDC
g3 <- umap_plot %>% ggplot(aes(x=V1, y=V2, color=Condition)) +
  geom_point(size=1.5) + xlab("UMAP1") + ylab("UMAP2") +
  geom_text(label=colnames(se_cvid_ctrl), nudge_y = 0.1,) +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("PBMC - Stimulated Samples")
```

```{r UMAPS_2, echo=FALSE}
combined_plot <- g1 + g3 + plot_layout(guides = "collect") & theme(legend.position = "bottom")

plot(combined_plot)
```

## DEGs (Deseq) Heat Maps

```{r lps_deseq, include=FALSE}
# format data into a DESeq2-friendly Data Structure
counts <- assay(se_cvid_LPS, "counts")
meta_data <- colData(se_cvid_LPS)
Condition <- meta_data$Condition

dds <- DESeqDataSetFromMatrix(countData=counts, colData=meta_data, design=~Condition)
#colData is a data frame of demographic/phenotypic data

# filter for genes with significant expression  
# expression should be > 0, but > 100 should be better
dds<-dds[rowSums(counts(dds))>1,] #Gene Filtering

# fit negative binomial regression
dds<-DESeq(dds) #Performs estimation of size factors, dispersion, and negative binomial GLM f#itting

# extracting results, ordering them by adjusted p-value
res <- results(dds)[order(results(dds)[,6]),]
#res[1:10,]

# store the name of the top 1000 most significant genes, per Dr. Johnson
lps_deseq <- data.frame(res)
lps_deseq1000 <- rownames(res)[1:1000]
```


```{r lps_deseq_HA, include=FALSE}
# Make a Heatmap of DEGs

# All together: extract the log cpm of the top 1000 genes and store that in a matrix
# # get the name of the genes ordered by adjusted p-val
top_genes <- rownames(results(dds)[order(results(dds)$padj),])[1:250]
# # convert log cpm to a matrix structure sorted by the top genes
mat <- as.matrix(assay(se_cvid_LPS,"log_counts_cpm"))[top_genes,] 

# normalize data row-wise
mat <- t(scale(t(mat)))

# store IDC into a single-column dataframe
df <- data.frame(Condition=colData(se_cvid_LPS)$Condition) 

# create HeatMap with some annotations
ha <- HeatmapAnnotation(df=df, col=list(Condition=c("Healthy"="Blue","Deficient"="Red")))
HM3 <- Heatmap(mat,
               name = "Log CPM",
               show_row_names=F,
               show_column_names=F,
               row_dend_side = "right",
               top_annotation=ha,
               column_title = "PBMC - Stimulated Samples",
               heatmap_legend_param = list(direction = "horizontal", 
                                           title_gp = gpar(fontsize = 8), 
                                           labels_gp = gpar(fontsize = 8)))

```

```{r deseq_Heatmaps_2}
h1 <- grid.grabExpr(draw(HM1,
                         heatmap_legend_side = "bottom",
                         annotation_legend_side = "bottom"))

h3 <- grid.grabExpr(draw(HM3,
                         heatmap_legend_side = "bottom",
                         annotation_legend_side = "bottom"))

# Arrange them separately
grid.arrange(h1, h3, ncol = 2)
```

## Combined DEGs (DeSeq)
```{r combined_results_2}
common_genes <- intersect(row.names(nasal_deseq), row.names(lps_deseq))

combined_results <- data.frame(gene = common_genes) %>%
    mutate(
        nasal_padj = nasal_deseq[common_genes,]$padj,
        lps_padj = lps_deseq[common_genes,]$padj,
        combined_chisq = -2*(log(nasal_padj) + log(lps_padj)),
        nasal_effect = nasal_deseq[common_genes,]$log2FoldChange,
        lps_effect = lps_deseq[common_genes,]$log2FoldChange,
    ) %>% 
    transmute(
        gene = gene,
        combined_p = 1 - pchisq(combined_chisq, 4),
        nasal_effect = ifelse(nasal_effect > 0, "Upregulated", "Downregulated"),
        lps_effect = ifelse(lps_effect > 0, "Upregulated", "Downregulated"),
        reg_agreement = ifelse(lps_effect == nasal_effect, "Same", "Opposite"),
    ) %>% 
    mutate(across(where(is.character), as.factor))%>% 
    filter(!is.na(combined_p)) %>%
    arrange(combined_p) %>% 
    slice_head(n = 250)

row.names(combined_results) <- combined_results$gene
combined_results$gene <- NULL

datatable(combined_results)
```

## Pathway Analysis - Any {.tabset}
```{r include = FALSE}
# based on the results of DEseq, cross-reference the gene with pathway databases to infer which pathways are the most significant

genes <- row.names(combined_results)
enriched <- enrichr(genes, dbs)
```

### DESeq2 WikiPathways

```{r}
datatable(enriched$WikiPathways_2016)
```

### DESeq2 KEGG

```{r}
datatable(enriched$KEGG_2016)
```

### DESeq2 Panther

```{r}
datatable(enriched$Panther_2016)
```

### DESeq2 Reactome

```{r}
datatable(enriched$Reactome_2016)
```

## Pathway Analysis - Same {.tabset}
```{r include=FALSE}
# based on the results of DEseq, cross-reference the gene with pathway databases to infer which pathways are the most significant

genes <- row.names(filter(combined_results, reg_agreement == "Same"))
enriched <- enrichr(genes, dbs)
```

### DESeq2 WikiPathways

```{r}
datatable(enriched$WikiPathways_2016)
```

### DESeq2 KEGG

```{r}
datatable(enriched$KEGG_2016)
```

### DESeq2 Panther

```{r}
datatable(enriched$Panther_2016)
```

### DESeq2 Reactome

```{r}
datatable(enriched$Reactome_2016)
```

## Pathway Analysis - Opposite {.tabset}
```{r include=FALSE}
# based on the results of DEseq, cross-reference the gene with pathway databases to infer which pathways are the most significant

genes <- row.names(filter(combined_results, reg_agreement == "Opposite"))
enriched <- enrichr(genes, dbs)
```
### DESeq2 WikiPathways

```{r}
datatable(enriched$WikiPathways_2016)
```

### DESeq2 KEGG

```{r}
datatable(enriched$KEGG_2016)
```

### DESeq2 Panther

```{r}
datatable(enriched$Panther_2016)
```

### DESeq2 Reactome

```{r}
datatable(enriched$Reactome_2016)
```
