---
title: "Comparative Results Analysis of Immune Deficiency RNA-Seq Data"
author: W. Evan Johnson, Ph.D., Ian Mc Farlane, Bram Stults
date: "`r Sys.Date()`"
header-includes:
   - \usepackage{amsmath}
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    theme: "flatly"
editor_options: 
  chunk_output_type: console
tables: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(readr)

```

##### Load top 250 DEGs per sample/group

```{r}
# WD set to /Deliverable 3/data directory
# setwd("~/Documents/.../Meta-omic-Analysis-Capstone/Deliverable 3")

results_1_df <- read_csv("data/NasalCVID_results_top250.csv")
nasal_deseq <- results_1_df$deseq250
nasal_limma <- results_1_df$limma250

results_2_df <- read_csv("data/PBMC_results_top250.csv")
ctrl_deseq <- results_2_df$ctrl_deseq
ctrl_limma <- results_2_df$ctrl_limma
LPS_deseq <- results_2_df$LPS_deseq
LPS_limma <- results_2_df$LPS_limma

```

## Comparison {.tabset}

### LPS v. Healthy Control {.tabset}

#### Common Significant genes

```{r}
(common_deseq <- intersect(LPS_deseq, ctrl_deseq))
(common_limma <- intersect(LPS_limma, ctrl_limma))
```

#### Significant genes for LPS group only

```{r}
setdiff(LPS_deseq, common_deseq)
setdiff(LPS_limma, common_limma)
```

#### Significant genes for Control group only

```{r}
setdiff(ctrl_deseq, common_deseq)
setdiff(ctrl_limma, common_limma)
```

#### Similarity Coefficients for Limma, DESeq analyses

```{r}
print("DESeq similarity coefficient")
length(common_deseq) / length(union(LPS_deseq, ctrl_deseq))

print("LIMMA similarity coefficient")
length(common_limma) / length(union(LPS_limma, ctrl_limma))
```

### Nasal and Control PBMC {.tabset}

#### Common Significant genes:

```{r}

(common_deseq <- intersect(nasal_deseq, ctrl_deseq)) 
(common_limma <- intersect(nasal_limma, ctrl_limma))

```

#### Significant genes for nasal mucusal samples only

```{r}
setdiff(nasal_deseq, common_deseq)
setdiff(nasal_limma, common_limma)
```

#### Significant genes for Control group only

```{r}
setdiff(ctrl_deseq, common_deseq)
setdiff(ctrl_limma, common_limma)
```

#### Similarity Coefficients for Nasal v. PBMC Limma, DESeq analyses

```{r}

print("DESeq similarity coefficient for Nasal v. PBMC samples")
length(common_deseq) / length(union(nasal_deseq, ctrl_deseq))

print("LIMMA similarity coefficient for Nasal v. PBMC samples")
length(common_limma) / length(union(nasal_limma, ctrl_limma))

```
