---
title: "Comparative Results Analysis of Immune Deficiency RNA-Seq Data"
author: Ian Mc Farlane, Bram Stults advised by W. Evan Johnson, Ph.D.
date: "`r Sys.Date()`"
header-includes:
   - \usepackage{amsmath}
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    theme: "flatly"
editor_options: 
  chunk_output_type: console
tables: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#load packages
if (!require("pacman")) install.packages("pacman")
pacman::p_load(here,
               tidyverse,
               magrittr,
               janitor,
               scales,
               ggrepel,
               DT)
library(readr)

```

### Load top 250 DEGs per sample/group

```{r}
# WD set to /Deliverable 3/data directory
# setwd("~/Documents/.../Meta-omic-Analysis-Capstone/Deliverable 6")

top.results.1.df <- read_csv("data/Nasopharynx_CVID_top250_results.csv")
nasophar.deseq <- top.results.1.df$deseq250
nasophar.limma <- top.results.1.df$limma250

#complete list of nasopharyngeal sample results
all.nasophar.res <- read_csv("data/Nasopharynx_CVID_all_results.csv")
names(all.nasophar.res)[names(all.nasophar.res) == "...1"] <- "genes"

top.results.2.df <- read_csv("data/PBMC_CVID_top250_results.csv")
ctrl.deseq <- top.results.2.df$ctrl_deseq
ctrl.limma <- top.results.2.df$ctrl_limma
LPS.deseq <- top.results.2.df$LPS_deseq
LPS.limma <- top.results.2.df$LPS_limma

#complete list of unstimulated PBMC sample results
all.pbmc.res <- read_csv("data/PBMC_CVID_all_results.csv")
names(all.pbmc.res)[names(all.pbmc.res) == "...1"] <- "genes"

```


## Comparison {.tabset}

## Nasopharyngeal and Healthy Control PBMC {.tabset}

### Common Significant genes:

```{r}

(common.deseq <- intersect(nasophar.deseq, ctrl.deseq)) 
(common.limma <- intersect(nasophar.limma, ctrl.limma))

```

### Significant genes for nasopharyngeal mucusal samples only

```{r}
setdiff(nasophar.deseq, common.deseq)
setdiff(nasophar.limma, common.limma)
```

### Significant genes for Control group only

```{r}
setdiff(ctrl.deseq, common.deseq)
setdiff(ctrl.limma, common.limma)
```

### Similarity Coefficients for Nasopharyngeal v. PBMC DESeq2 analysis

```{r}

print("DESeq similarity coefficient for Nasopharyngeal v. PBMC samples")
length(common.deseq) / length(union(nasophar.deseq, ctrl.deseq))

#storing DESeq2 coefficient for Monte Carlo verification
jacc.index <- length(common.deseq) / length(union(nasophar.deseq, ctrl.deseq))

print("LIMMA similarity coefficient for Nasopharyngeal v. PBMC samples")
length(common.limma) / length(union(nasophar.limma, ctrl.limma))

```

## Combined DEGs (DeSeq)
```{r}
nasal.res <- all.nasophar.res
blood.res <- all.pbmc.res

common_genes <- intersect(nasal.res$genes,blood.res$genes)
significance.threshold <- 0.2

merged.res <- tibble(genes = common_genes) %>% 
    merge(nasal.res, by="genes") %>% 
    merge(blood.res, by="genes", suffixes=c(".nasal", ".blood"))

if ( nrow(merged.res) != length(common_genes) ) {
    stop(paste(
        "There was an error when merging the datasets.", 
        "The number of rows (", nrow(merged.res), ")",
        "doesn't match with the number of common genes",
        "(", length(common_genes), ")."
    ))
}

######## 
# NOTE: once the baseline of PBMC gets fixed, change < to > in blood.regulation
######## 

merged.res <- merged.res %>% 
    mutate(
        blood.padj = padj.blood, 
        nasal.padj = padj.nasal,
        combined.chisq = -2*(log(blood.padj) + log(nasal.padj)),
        combined.p = 1 - pchisq(combined.chisq, 4),
        blood.regulation = ifelse(log2FoldChange.blood < 0, "Upregulated", "Downregulated"),
        nasal.regulation = ifelse(log2FoldChange.nasal > 0, "Upregulated", "Downregulated"),
        regulation.agreement = ifelse(blood.regulation == nasal.regulation, "Same", "Opposite")
    ) %>% 
    mutate(across(where(is.character), as.factor)) %>% 
    filter(!is.na(combined.p)) %>%
    arrange(combined.p)

any.sig.res <- merged.res %>% 
    mutate(blood.padj = padj.blood, nasal.padj = padj.nasal) %>%
    dplyr::select(
        genes, combined.p, blood.padj, nasal.padj,
        blood.regulation, nasal.regulation, regulation.agreement
    ) %>%
    slice_head(n = 250)

mutual.sig.res <- merged.res %>% 
    mutate(blood.padj = padj.blood, nasal.padj = padj.nasal) %>%
    dplyr::select(
        genes, combined.p, blood.padj, nasal.padj,
        blood.regulation, nasal.regulation, regulation.agreement
    ) %>%
    filter(blood.padj < significance.threshold, nasal.padj < significance.threshold)


datatable(mutual.sig.res)
```

## Agreement of regulation

### Between mutually significant genes
```{r}
mean(mutual.sig.res$regulation.agreement == "Same")
```

### Between Fischer-significant genes
```{r}
mean(any.sig.res$regulation.agreement == "Same")
```

## Contingency tables of of regulation directions

### Between mutually significant genes
```{r}
(mutual.contingency <- table(dplyr::select(mutual.sig.res, blood.regulation, nasal.regulation)))

chisq.test(mutual.contingency)
fisher.test(mutual.contingency)
```

### Between Fischer-significant genes
```{r}
(any.contingency <- table(dplyr::select(any.sig.res, blood.regulation, nasal.regulation)))

chisq.test(any.contingency)
fisher.test(any.contingency)
```

## Monte Carlo Simulation of Jaccard Index - Result verification

```{r}

# function for commonality coefficient 
commonality.coefficient <- function(sample1, sample2) {
  intersection <- intersect(sample1$genes, sample2$genes)
  union <- union(sample1$genes, sample2$genes)
  return(length(intersection) / length(union))
}

# number of simulations
num.sims <- 10#0000

# number of genes to sample
sample.size <- 250

# result vector
common.coeffs <- numeric(num.sims)

# Loop
for (i in 1:num.sims) {
  
  # randomly selecting 250 row indices from `all.pbmc.res` (complete results)
  blood.indices <- sample(nrow(all.pbmc.res), 
                          size = sample.size, 
                          replace = F) 
  blood.sample <- all.pbmc.res[blood.indices, ]
  
  # randomly selecting 250 row indices from `all.nasophar.res` (complete results)
  nasophar.indices <- sample(nrow(all.nasophar.res), 
                      size = sample.size, 
                      replace = F)  
  nasophar.sample <- all.nasophar.res[nasophar.indices, ]

  # calculating and storing commonality coefficient for samples
  coefficient <- commonality.coefficient(blood.sample, nasophar.sample)
  common.coeffs[i] <- coefficient
}

mean.commonality <- mean(common.coeffs)
sd.commonality <- sd(common.coeffs)

cat("Mean Commonality Coefficient:", mean.commonality, "\n")
cat("Standard deviation:", sd.commonality, "\n")

hist(common.coeffs, 
     main = "Distribution of Commonality Coefficients", 
     xlab = "Commonality Coefficient")

#Probability of obtaining Jaccard index greater than obtained DESeq2 coefficient
mean(common.coeffs >= jacc.index) #mean of True/False logical vector T/F --> 1/0



```
If the 250 genes in each list were completely random, there is a 2.988% chance that the resulting commonality coefficients of those lists would be as large as or larger than 0.008064516 obtained here.


## Volcano Plot - PBMC
```{r}

all.pbmc.res <- all.pbmc.res |>
  mutate(gene_type=case_when(log2FoldChange >= 1 & padj <= 0.05 ~"up",
                             log2FoldChange <= -1 & padj <= 0.05 ~"down",
                             TRUE ~ "ns" ))
all.pbmc.res |> 
  count(gene_type)

all.pbmc.res |>
  distinct(gene_type) |> 
  pull()

cols <- c("up"="slategray2", "down"="orange2", "ns"="grey50")
sizes <- c("up"=2, "down"=2, "ns"=1)
alphas <- c("up"=1, "down"=1, "ns"=0.5)

all.pbmc.res |> 
  ggplot(aes(x=log2FoldChange,
             y=-log10(padj),
             fill=gene_type,
             size=gene_type,
             alpha=gene_type)) +
  geom_point(shape=21,
             color="black") +
  geom_vline(xintercept=c(log2(0.5), log2(2)),
          linetype="dashed")+
  geom_hline(yintercept=-log10(0.05),
             linetype="dashed")+
  scale_fill_manual(values=cols)+
  scale_size_manual(values=sizes)+
  scale_alpha_manual(values=alphas)+
  scale_x_continuous(breaks=c(seq(-10,10,2)),
                     limits=c(-10,10)
                     )
  
```


## Volcano Plot - Nasopharynx
```{r}

all.nasophar.res <- all.nasophar.res |>
  mutate(gene_type=case_when(log2FoldChange >= 1 & padj <= 0.05 ~"up",
                             log2FoldChange <= -1 & padj <= 0.05 ~"down",
                             TRUE ~ "ns" ))
all.nasophar.res |> 
  count(gene_type)

all.nasophar.res |>
  distinct(gene_type) |> 
  pull()

cols <- c("up"="slategray2", "down"="orange2", "ns"="grey50")
sizes <- c("up"=2, "down"=2, "ns"=1)
alphas <- c("up"=1, "down"=1, "ns"=0.5)

all.nasophar.res |> 
  ggplot(aes(x=log2FoldChange,
             y=-log10(padj),
             fill=gene_type,
             size=gene_type,
             alpha=gene_type)) +
  geom_point(shape=21,
             color="black") +
  geom_vline(xintercept=c(log2(0.5), log2(2)),
          linetype="dashed")+
  geom_hline(yintercept=-log10(0.05),
             linetype="dashed")+
  scale_fill_manual(values=cols)+
  scale_size_manual(values=sizes)+
  scale_alpha_manual(values=alphas)+
  scale_x_continuous(breaks=c(seq(-10,10,2)),
                     limits=c(-10,10)
                     )

  
```



