---
title: "Comparative Results Analysis of Immune Deficiency RNA-Seq Data"
author: W. Evan Johnson, Ph.D., Ian Mc Farlane, Bram Stults
date: "`r Sys.Date()`"
header-includes:
   - \usepackage{amsmath}
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    theme: "flatly"
editor_options: 
  chunk_output_type: console
tables: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(Rsubread)
library(Rsamtools)
library(SummarizedExperiment)
library(edgeR)
library(DESeq2)
library(ComplexHeatmap)
library(TBSignatureProfiler)
library(umap)
library(tidyverse)
library(DT)
library(GSVA)
library(enrichR)
library(biomaRt)
library(rprojroot)
library(fs)
library(dplyr)

# define pathway libraries
dbs <- c("WikiPathways_2016","KEGG_2016","Panther_2016","Reactome_2016")

showPathways <- function(gene_names) {
    
    if (length(gene_names) < 10) {
        print(paste(length(gene_names), "genes are not enough"))
        return()
    }
    
    invisible(capture.output(
      enriched <- enrichr(gene_names, dbs)
    ))
    
    combined_df <- bind_rows(
          enriched$WikiPathways_2016 %>% mutate(database = "WikiPathways"),
          enriched$KEGG_2016 %>% mutate(database = "KEGG"),
          enriched$Panther_2016 %>% mutate(database = "Panther"),
          enriched$Reactome_2016 %>% mutate(database = "Reactome")
        ) %>%
        arrange(Adjusted.P.value)
    
    combined_df$Adjusted.P.value <- NULL
    combined_df$Old.Adjusted.P.value <- NULL
    
    # Display the combined data frame
    datatable(combined_df)
}

```

### Load top 250 DEGs per sample/group

```{r}
# WD set to /Deliverable 3/data directory
# setwd("~/Documents/.../Meta-omic-Analysis-Capstone/Deliverable 3")

results_1_df <- read_csv("data/NasalCVID_results_top250.csv")
nasophar.deseq <- results_1_df$deseq250
nasophar.limma <- results_1_df$limma250

results_2_df <- read_csv("data/PBMC_results_top250.csv")
ctrl.deseq <- results_2_df$ctrl_deseq
ctrl.limma <- results_2_df$ctrl_limma
LPS.deseq <- results_2_df$LPS_deseq
LPS.limma <- results_2_df$LPS_limma

results_3_df <- read_csv("data/TB_results_top250.csv")
nasal.deseq <- results_3_df$nasal.deseq
nasal.limma <- results_3_df$nasal.limma
blood.deseq <- results_3_df$blood.deseq
blood.limma <- results_3_df$blood.limma



```

## Comparison {.tabset}

## PBMC - LPS v. Healthy Control {.tabset}

### Common Significant genes

```{r}
(common.deseq <- intersect(LPS.deseq, ctrl.deseq))
(common.limma <- intersect(LPS.limma, ctrl.limma))
```

### Significant genes for LPS group only

```{r}
setdiff(LPS.deseq, common.deseq)
setdiff(LPS.limma, common.limma)
```

### Significant genes for Control group only

```{r}
setdiff(ctrl.deseq, common.deseq)
setdiff(ctrl.limma, common.limma)
```

### Similarity Coefficients for Limma, DESeq analyses

```{r}
print("DESeq similarity coefficient")
length(common.deseq) / length(union(LPS.deseq, ctrl.deseq))

print("LIMMA similarity coefficient")
length(common.limma) / length(union(LPS.limma, ctrl.limma))
```

## Nasopharynx and Healthy Control PBMC {.tabset}

### Common Significant genes:

```{r}

(common.deseq <- intersect(nasal.deseq, ctrl.deseq)) 
(common.limma <- intersect(nasal.limma, ctrl.limma))

```

### Significant genes for nasal mucusal samples only

```{r}
setdiff(nasal.deseq, common.deseq)
setdiff(nasal.limma, common.limma)
```

### Significant genes for Control group only

```{r}
setdiff(ctrl.deseq, common.deseq)
setdiff(ctrl.limma, common.limma)
```

### Similarity Coefficients for Nasal v. PBMC Limma, DESeq analyses

```{r}

print("DESeq similarity coefficient for Nasal v. PBMC samples")
length(common.deseq) / length(union(nasal.deseq, ctrl.deseq))

print("LIMMA similarity coefficient for Nasal v. PBMC samples")
length(common.limma) / length(union(nasal.limma, ctrl.limma))

```




## Nasopharynx and LPS Treatment PBMC {.tabset}

### Common Significant genes:

```{r}

(common.deseq <- intersect(nasal.deseq, LPS.deseq)) 
(common.limma <- intersect(nasal.limma, LPS.limma))

```

### Significant genes for nasal mucusal samples only

```{r}
setdiff(nasal.deseq, common.deseq)
setdiff(nasal.limma, common.limma)
```

### Significant genes for LPS Treatment group only

```{r}
setdiff(LPS.deseq, common.deseq)
setdiff(LPS.limma, common.limma)
```

### Similarity Coefficients for Nasal v. PBMC Limma, DESeq analyses

```{r}

print("DESeq similarity coefficient for Nasal v. PBMC samples")
length(common.deseq) / length(union(nasal.deseq, LPS.deseq))

print("LIMMA similarity coefficient for Nasal v. PBMC samples")
length(common.limma) / length(union(nasal.limma, LPS.limma))

```

# Deliverable 5

## TB Nasal and CVID Nasopharynx Samples {.tabset}

### Common Significant genes:

```{r}

( common.deseq <- intersect(nasophar.deseq, nasal.deseq) ) 
( common.limma <- intersect(nasophar.limma, nasal.limma) )

```

### Significant genes for CVID nasopharynx mucosal samples only

```{r}
setdiff(nasophar.deseq, common.deseq)
setdiff(nasophar.limma, common.limma)
```

### Significant genes for TB nasal sample group only

```{r}
setdiff(nasal.deseq, common.deseq)
setdiff(nasal.limma, common.limma)
```

### Similarity Coefficients for CVID Nasopharynx v. TB Nasal Limma, DESeq analyses

```{r}

print("DESeq similarity coefficient for CVID Nasopharynx v. TB Nasal samples")
length(common.deseq) / length(union(nasophar.deseq, nasal.deseq))

print("LIMMA similarity coefficient for CVID Nasopharynx v. TB Nasal samples")
length(common.limma) / length(union(nasophar.limma, nasal.limma))

```

### Pathways {.tabset}

#### Deseq
```{r}
showPathways(common.deseq)
```


#### LIMMA
```{r}
showPathways(common.limma)
```

## TB Blood and CVID PBMC LPS Treatment Samples {.tabset}

### Common Significant genes:

```{r}

(common.deseq <- intersect(blood.deseq, LPS.deseq)) 
(common.limma <- intersect(blood.limma, LPS.limma))

```

### Significant genes for TB samples only

```{r}
setdiff(blood.deseq, common.deseq)
setdiff(blood.limma, common.limma)
```

### Significant genes for PBMC LPS Treatment group only

```{r}
setdiff(LPS.deseq, common.deseq)
setdiff(LPS.limma, common.limma)
```

### Similarity Coefficients for TB Blood v.CVID PBMC Limma, DESeq analyses

```{r}

print("DESeq similarity coefficient for TB Blood v. CVID PBMC samples")
length(common.deseq) / length(union(blood.deseq, LPS.deseq))

print("LIMMA similarity coefficient for Nasal v. PBMC samples")
length(common.limma) / length(union(blood.limma, LPS.limma))

```

### Pathways {.tabset}

#### Deseq
```{r}
showPathways(common.deseq)
```


#### LIMMA
```{r}
showPathways(common.limma)
```

## TB Blood and CVID PBMC Non-Treatment (LPS control) Samples {.tabset}

### Common Significant genes:

```{r}

(common.deseq <- intersect(blood.deseq, ctrl.deseq)) 
(common.limma <- intersect(blood.limma, ctrl.limma))

```

### Significant genes for TB Blood samples only

```{r}
setdiff(blood.deseq, common.deseq)
setdiff(blood.limma, common.limma)
```

### Significant genes for PBMC Non-treatment group only

```{r}
setdiff(ctrl.deseq, common.deseq)
setdiff(ctrl.limma, common.limma)
```

### Similarity Coefficients for TB Blood v.CVID PBMC Limma, DESeq analyses

```{r}

print("DESeq similarity coefficient for TB Blood v. CVID PBMC samples")
length(common.deseq) / length(union(blood.deseq, ctrl.deseq))

print("LIMMA similarity coefficient for Nasal v. PBMC samples")
length(common.limma) / length(union(blood.limma, ctrl.limma))
```

### Pathways {.tabset}

#### Deseq
```{r}
showPathways(common.deseq)
```


#### LIMMA
```{r}
showPathways(common.limma)
```

## TB Blood and CVID Nasopharynx Samples {.tabset}

### Common Significant genes:

```{r}

(common.deseq <- intersect(blood.deseq, nasophar.deseq)) 
(common.limma <- intersect(blood.limma, nasophar.limma))

```

### Significant genes for TB Blood samples only

```{r}
setdiff(blood.deseq, common.deseq)
setdiff(blood.limma, common.limma)
```

### Significant genes for PBMC Non-treatment group only

```{r}
setdiff(nasophar.deseq, common.deseq)
setdiff(nasophar.limma, common.limma)
```

### Similarity Coefficients for TB Blood v. CVID PBMC Limma, DESeq analyses

```{r}

print("DESeq similarity coefficient for TB Blood v. CVID PBMC samples")
length(common.deseq) / length(union(blood.deseq, nasophar.deseq))

print("LIMMA similarity coefficient for Nasal v. PBMC samples")
length(common.limma) / length(union(blood.limma, nasophar.limma))
```

### Pathways {.tabset}

#### Deseq
```{r}
showPathways(common.deseq)
```


#### LIMMA
```{r}
showPathways(common.limma)
```

## TB Nasal and CVID PBMC LPS-Treatment Samples {.tabset}

### Common Significant genes:

```{r}

(common.deseq <- intersect(nasal.deseq, LPS.deseq)) 
(common.limma <- intersect(nasal.limma, LPS.limma))

```

### Significant genes for TB Blood samples only

```{r}
setdiff(nasal.deseq, common.deseq)
setdiff(nasal.limma, common.limma)
```

### Significant genes for PBMC Non-treatment group only

```{r}
setdiff(LPS.deseq, common.deseq)
setdiff(LPS.limma, common.limma)
```

### Similarity Coefficients for TB Blood v.CVID PBMC Limma, DESeq analyses

```{r}

print("DESeq similarity coefficient for TB Blood v. CVID PBMC samples")
length(common.deseq) / length(union(nasal.deseq, LPS.deseq))

print("LIMMA similarity coefficient for Nasal v. PBMC samples")
length(common.limma) / length(union(nasal.limma, LPS.limma))
```

### Pathways {.tabset}

#### Deseq
```{r}
showPathways(common.deseq)
```


#### LIMMA
```{r}
showPathways(common.limma)
```

## TB Nasal and CVID PBMC Non-Treatment (LPS control) Samples {.tabset}

### Common Significant genes:

```{r}

(common.deseq <- intersect(nasal.deseq, ctrl.deseq)) 
(common.limma <- intersect(nasal.limma, ctrl.limma))

```

### Significant genes for TB Blood samples only

```{r}
setdiff(nasal.deseq, common.deseq)
setdiff(nasal.limma, common.limma)
```

### Significant genes for PBMC Non-treatment group only

```{r}
setdiff(ctrl.deseq, common.deseq)
setdiff(ctrl.limma, common.limma)
```

### Similarity Coefficients for TB Blood v.CVID PBMC Limma, DESeq analyses

```{r}

print("DESeq similarity coefficient for TB Blood v. CVID PBMC samples")
length(common.deseq) / length(union(nasal.deseq, ctrl.deseq))

print("LIMMA similarity coefficient for Nasal v. PBMC samples")
length(common.limma) / length(union(nasal.limma, ctrl.limma))
```

### Pathways {.tabset}

#### Deseq
```{r}
showPathways(common.deseq)
```


#### LIMMA
```{r}
showPathways(common.limma)
```