---
title: "Integrative Analysis of\nCommon Variable Immunodeficiency (CVID)"
author: "Ian Mc Farlane, Bram Stults, advised by W. Evan Johnson, Ph.D."
date: "`r Sys.Date()`"
output: slidy_presentation
header-includes:
   - \usepackage{amsmath}
editor_options: 
  chunk_output_type: console
tables: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(readr)
library(Rsubread)
library(Rsamtools)
library(SummarizedExperiment)
library(edgeR)
library(DESeq2)
library(ComplexHeatmap)
library(TBSignatureProfiler)
library(umap)
library(tidyverse)
library(DT)
library(GSVA)
library(enrichR)
library(biomaRt)
library(rprojroot)
library(fs)
library(dplyr)
library(patchwork)
library(gridExtra)
library(ComplexHeatmap)

# define pathway libraries
dbs <- c("WikiPathways_2016","KEGG_2016","Panther_2016","Reactome_2016")

```


```{r read_in_blood, include=FALSE}
# read RNA expression counts and patient data
load(file="data/readMatrix_PBMC_BulkRNAseq_CVID.Rdata")

mt.counts.df <- read.csv("data/translated_genes.txt", row.names = 1)
groups <- sample.info$disease 

subs = c(UCVID = "Deficient (CVID)",
         CVID = "Deficient (CVID)",
         healthy = "Control (Healthy)",
         NCVID = "remove")


groups <- unname( subs[match(groups, names(subs))] )

# store index of NCVID subjects
remove_NCVID_index <- which(groups == "remove")

# update group label vector
groups <- groups[groups != "remove"] 
groups <- factor(groups)


#updating dataframes
mt.counts.df <- mt.counts.df[,-remove_NCVID_index]
sample.info <- sample.info[-remove_NCVID_index,]
sample.info$Condition <- groups

#  re-store results
counts <- mt.counts.df
meta_data <- sample.info

# isolate group indicator variable
group <- meta_data$disease

# aggregate patient and expression data into a single object
se_cvid <- SummarizedExperiment(assays=list(counts=counts),
                     colData = meta_data)


# isolating treatment group experiments
treatments <- meta_data$treatment
ctrl_index <- which(treatments == "ctrl")
LPS_index <- which(treatments == "LPS")

counts_ctrl <- counts[,-LPS_index]
counts_LPS <- counts[,-ctrl_index]

meta_data_ctrl <- meta_data[-LPS_index,]
meta_data_LPS <- meta_data[-ctrl_index,]

se_cvid_ctrl <- SummarizedExperiment(assays=list(counts=counts_ctrl),
                     colData = meta_data_ctrl)
se_cvid_LPS <- SummarizedExperiment(assays=list(counts=counts_LPS),
                     colData = meta_data_LPS)

# create counts per million, log counts and log counts per million features
# and save them into new assays
se_cvid <- mkAssay(se_cvid, log = TRUE, counts_to_CPM = TRUE)
# same for control only
se_cvid_ctrl <- mkAssay(se_cvid_ctrl, log = TRUE, counts_to_CPM = TRUE)
# same for LPS treatment group only
se_cvid_LPS <- mkAssay(se_cvid_LPS, log = TRUE, counts_to_CPM = TRUE)
```


```{r read_in_nasal, include=FALSE}

# read RNA expression counts
counts <- read.table("data/features_combined.txt", sep="\t", 
                     header=T, row.names=1)
# read patient data
meta_data <- read.table("data/meta_data.txt", sep="\t", 
                     header=T, row.names=1)

# isolate group indicator variable
group <- meta_data$Condition
group <- ifelse(group == "Control", "Control (Healthy)", "Deficient (CVID)")
meta_data$Condition <- group

# aggregate patient and expression data into a single object
se_immunedef <- SummarizedExperiment(assays=list(counts=counts),
                     colData = meta_data)

# create counts per million, log counts and log counts per million features 
# and save them into new assays.
se_immunedef <- mkAssay(se_immunedef, log = TRUE, counts_to_CPM = TRUE)

# display all assays
assays(se_immunedef)
```

# Review

## Methods

- We analyzed:
  - Nasopharyngeal Samples from Dr Johnson
  - PBMC Samples with LPS stimulated and control groups from Dr Maglione.

- We compared the results of the first analysis to the results of each of the groups of the second analysis:
  - Nasopharyngeal vs LPS stimulated
  - Nasopharyngeal vs Control (LPS UNstimulated)
    
## Results summary

- We identified the following types of pathways to be significantly affected in both comparisons:
  - Focal adhesion
  - Inflammation
  - Chemokine receptors

- Overarching Genes:
  - ITGA4: Encodes integrin alpha-4, a cell adhesion molecule involved in leukocyte trafficking and immune cell migration.
  - GPR55: Encodes a G protein-coupled receptor implicated in inflammatory responses, pain signaling, and cancer progression.
  - CCR4: Encodes a chemokine receptor that regulates immune cell trafficking, particularly T-cell migration to inflammation sites.


<!-- Fitting models -->
```{r nasal_deseq_0, include=FALSE, cache=TRUE}
# format data into a DESeq2-friendly Data Structure
counts <- assay(se_immunedef, "counts")
meta_data <- colData(se_immunedef)
Condition <- meta_data$Condition

dds <- DESeqDataSetFromMatrix(countData = counts, colData=meta_data, design=~Condition)
#colData is a data frame of demographic/phenotypic data

# filter for genes with significant expression  
# expression should be > 0, but > 100 should be better
dds<-dds[rowSums(counts(dds))>1,] #Gene Filtering

# fit negative binomial regression
dds<-DESeq(dds) #Performs estimation of size factors,dispersion, and negative binomial GLM f#itting

# extracting results, ordering them by adjusted p-value
res <- results(dds)[order(results(dds)[,6]),]

# store the name of the top 1000 most significant genes, per Dr. Johnson
nasal_deseq <- data.frame(res)
nasal_deseq1000 <- rownames(res)[1:1000]
```

```{r ctrl_deseq_0, include=FALSE, cache=TRUE}
# format data into a DESeq2-friendly Data Structure
counts <- assay(se_cvid_ctrl, "counts")
meta_data <- colData(se_cvid_ctrl)
Condition <- meta_data$Condition

dds <- DESeqDataSetFromMatrix(countData=counts, colData=meta_data, design=~Condition)
#colData is a data frame of demographic/phenotypic data

# filter for genes with significant expression  
# expression should be > 0, but > 100 should be better
dds<-dds[rowSums(counts(dds))>1,] #Gene Filtering

# fit negative binomial regression
dds<-DESeq(dds) #Performs estimation of size factors, dispersion, and negative binomial GLM f#itting

# extracting results, ordering them by adjusted p-value
res <- results(dds)[order(results(dds)[,6]),]
#res[1:10,]

# store the name of the top 1000 most significant genes, per Dr. Johnson
ctrl_deseq <- data.frame(res)
ctrl_deseq1000 <- rownames(res)[1:1000]
```

```{r pbmc_control_results}
#top 1000 list used, pared to 250

ctrl_results <- data.frame(gene = ctrl_deseq1000) %>%
    mutate(
        ctrl_padj = ctrl_deseq[ctrl_deseq1000,]$padj,
        ctrl_effect = ctrl_deseq[ctrl_deseq1000,]$log2FoldChange
    ) %>% 
    mutate(
        ctrl_effect = ifelse(ctrl_effect > 0, "Upregulated", "Downregulated")
    ) %>% 
    mutate(across(where(is.character), as.factor)) %>% 
    arrange(ctrl_padj) %>% 
    slice_head(n = 250)

row.names(ctrl_results) <- ctrl_results$gene
ctrl_results$gene <- NULL
```


```{r lps_deseq_0, include=FALSE, cache=TRUE}
# format data into a DESeq2-friendly Data Structure
counts <- assay(se_cvid_LPS, "counts")
meta_data <- colData(se_cvid_LPS)
Condition <- meta_data$Condition

dds <- DESeqDataSetFromMatrix(countData=counts, colData=meta_data, design=~Condition)
#colData is a data frame of demographic/phenotypic data

# filter for genes with significant expression  
# expression should be > 0, but > 100 should be better
dds<-dds[rowSums(counts(dds))>1,] #Gene Filtering

# fit negative binomial regression
dds<-DESeq(dds) #Performs estimation of size factors, dispersion, and negative binomial GLM f#itting

# extracting results, ordering them by adjusted p-value
res <- results(dds)[order(results(dds)[,6]),]
#res[1:10,]

# store the name of the top 1000 most significant genes, per Dr. Johnson
lps_deseq <- data.frame(res)
lps_deseq1000 <- rownames(res)[1:1000]
```


```{r pbmc_lps_results}
#top 1000 list used, pared to 250

lps_results <- data.frame(gene = lps_deseq1000) %>%
    mutate(
        lps_padj = lps_deseq[lps_deseq1000,]$padj,
        lps_effect = lps_deseq[lps_deseq1000,]$log2FoldChange
    ) %>% 
    mutate(
        lps_effect = ifelse(lps_effect > 0, "Upregulated", "Downregulated")
    ) %>% 
    mutate(across(where(is.character), as.factor)) %>% 
    arrange(lps_padj) %>% 
    slice_head(n = 250)

row.names(lps_results) <- lps_results$gene
lps_results$gene <- NULL
```
<!-- END Fitting models -->

## Nasal VS Control (Unstimulated) - Combined DEGs (DeSeq)
```{r combined_results}
common_genes <- intersect(row.names(nasal_deseq), row.names(ctrl_deseq))
significance_threshold <- 0.2

combined_results <- data.frame(gene = common_genes) %>%
    mutate(
        nasal_padj = nasal_deseq[common_genes,]$padj,
        ctrl_padj = ctrl_deseq[common_genes,]$padj,
        combined_chisq = -2*(log(nasal_padj) + log(ctrl_padj)),
        nasal_effect = nasal_deseq[common_genes,]$log2FoldChange,
        ctrl_effect = ctrl_deseq[common_genes,]$log2FoldChange,
    ) %>% 
    transmute(
        gene = gene, nasal_padj = nasal_padj, ctrl_padj = ctrl_padj,
        combined_p = 1 - pchisq(combined_chisq, 4),
        nasal_effect = ifelse(nasal_effect > 0, "Upregulated", "Downregulated"),
        ctrl_effect = ifelse(ctrl_effect > 0, "Upregulated", "Downregulated"),
        reg_agreement = ifelse(ctrl_effect == nasal_effect, "Same", "Opposite"),
    ) %>% 
    mutate(across(where(is.character), as.factor))%>% 
    filter(
        !is.na(combined_p),
        nasal_padj < significance_threshold,
        ctrl_padj < significance_threshold,
    ) %>%
    arrange(combined_p) %>% 
    slice_head(n = 250)

row.names(combined_results) <- combined_results$gene
combined_results$gene <- NULL
combined_results$nasal_padj <- NULL
combined_results$ctrl_padj <- NULL


datatable(combined_results)
```

## Nasal VS Control (Unstimulated) - Combined Pathways
```{r}
# based on the results of DEseq, cross-reference the gene with pathway databases to infer which pathways are the most significant

showPathways <- function(gene_names) {
    invisible(capture.output(
      enriched <- enrichr(gene_names, dbs)
    ))
    
    combined_df <- bind_rows(
          enriched$WikiPathways_2016 %>% mutate(database = "WikiPathways"),
          enriched$KEGG_2016 %>% mutate(database = "KEGG"),
          enriched$Panther_2016 %>% mutate(database = "Panther"),
          enriched$Reactome_2016 %>% mutate(database = "Reactome")
        ) %>%
        arrange(Adjusted.P.value)
    
    combined_df$Adjusted.P.value <- NULL
    combined_df$Old.Adjusted.P.value <- NULL
    
    # Display the combined data frame
    datatable(combined_df)
}

showPathways(row.names(combined_results))
```



## Nasal VS LPS Stimulated - Combined DEGs (DeSeq)

```{r combined_results_2}
common_genes <- intersect(row.names(nasal_deseq), row.names(lps_deseq))

combined_results <- data.frame(gene = common_genes) %>%
    mutate(
        nasal_padj = nasal_deseq[common_genes,]$padj,
        lps_padj = lps_deseq[common_genes,]$padj,
        combined_chisq = -2*(log(nasal_padj) + log(lps_padj)),
        nasal_effect = nasal_deseq[common_genes,]$log2FoldChange,
        lps_effect = lps_deseq[common_genes,]$log2FoldChange,
    ) %>% 
    transmute(
        gene = gene, nasal_padj = nasal_padj, lps_padj = lps_padj,
        combined_p = 1 - pchisq(combined_chisq, 4),
        nasal_effect = ifelse(nasal_effect > 0, "Upregulated", "Downregulated"),
        lps_effect = ifelse(lps_effect > 0, "Upregulated", "Downregulated"),
        reg_agreement = ifelse(lps_effect == nasal_effect, "Same", "Opposite"),
    ) %>% 
    mutate(across(where(is.character), as.factor))%>% 
    filter(
        !is.na(combined_p),
        nasal_padj < significance_threshold,
        lps_padj < significance_threshold,
    ) %>%
    arrange(combined_p) %>% 
    slice_head(n = 250)

row.names(combined_results) <- combined_results$gene
combined_results$gene <- NULL
combined_results$nasal_padj <- NULL
combined_results$lps_padj <- NULL

datatable(combined_results)
```

## Nasal VS LPS Stimulated - Combined Pathways
```{r}
showPathways(row.names(combined_results))
```

# Update

## Methods
- We replicated this sort of analysis with Tuberculosis data provided by Dr Johnson
    - Blood and Nasal groups.
    
- We compared the results from the Blood and Nasal groups using similar techniques.

- We tried to compare the results to the previous analysis once more, but the DEG overlap was not significant enough to perform any meaningful analisys.


<!-- TB pipeline -->
```{r include=FALSE}
# read in and format patient data for both blood and nasal samples
meta_data <- read.csv("data/full_metadata.csv", row.names = 1)
meta_data$sample_name_rnacounts_file <- toupper(meta_data$sample_name_rnacounts_file)
meta_data$disease <- meta_data$status

# read expression counts
blood.counts.df <- read.csv("data/bloodcounts.csv", row.names = 1)
nasal.counts.df <- read.csv("data/nasalcounts.csv", row.names = 1)

# filter patient data without a match on expression count dataframes
relevant.rows <- meta_data$sample_name_rnacounts_file %in% c(names(blood.counts.df), names(nasal.counts.df)) 
meta_data <- meta_data[relevant.rows,]
meta_data$status <- factor(meta_data$status,levels = c("control","case"))

# split patient data into blood and nasal samples
blood.meta_data <- meta_data[meta_data$type == "blood",]
nasal.meta_data <- meta_data[meta_data$type == "nasal",]

# identify patients that performed both trials
samples_of_interest <- intersect(blood.meta_data$sample_id, nasal.meta_data$sample_id)

# blood: filter for the patients that also participated in nasal sampling
blood.mask <- blood.meta_data$sample_id %in% samples_of_interest
blood.meta_data <- blood.meta_data[blood.mask,]
columns <- blood.meta_data$sample_name_rnacounts_file
blood.counts.df <- blood.counts.df[,columns]

# nasal: filter for the patients that also participated in blood sampling
nasal.mask <- nasal.meta_data$sample_id %in% samples_of_interest
nasal.meta_data <- nasal.meta_data[nasal.mask,]
columns <- nasal.meta_data$sample_name_rnacounts_file
nasal.counts.df <- nasal.counts.df[,columns]


# aggregate patient and expression data into a single object for each set of samples
se.blood <- SummarizedExperiment(assays=list(counts=blood.counts.df), colData=blood.meta_data)
se.nasal <- SummarizedExperiment(assays=list(counts=nasal.counts.df), colData=nasal.meta_data)

# add log*cpm assays
se.blood <- mkAssay(se.blood, log = TRUE, counts_to_CPM = TRUE)
se.nasal <- mkAssay(se.nasal, log = TRUE, counts_to_CPM = TRUE)
```

```{r blood_vars, include=FALSE}
counts <- assay(se.blood, "counts")
meta_data <- colData(se.blood)
disease <- meta_data$status
group <- meta_data$status
x <- meta_data$sample_id
labels <- sapply(strsplit(x, "-"), function(parts) {
  paste0(as.numeric(parts[2]), "-", as.numeric(parts[3]))
})
```

```{r include=FALSE}
# fit UMAP model to log cpm data
# note: transpose data to make it tidy
set.seed(1)
umap.defaults$n_neighbors=5
umap_out <- umap(t(assay(se.blood,"log_counts_cpm")), config=umap.defaults)

# define dataframe with UMAP dimensions and IDC
umap_plot <- as.data.frame(umap_out$layout) # X, Y, ...
umap_plot$disease <- as.factor(disease) # color / group

# plot UMAP1 and UMAP2 by IDC
G1 <- umap_plot %>% ggplot(aes(x=V1, y=V2, color=disease)) +
  geom_point(size=1.5) + xlab("UMAP1") + ylab("UMAP2") +
  geom_text(label=labels, nudge_y = 0.1,) +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("UMAP Plot - Blood Treatment")
```

```{r message=FALSE, include=FALSE, cache=TRUE}
# format data into a DESeq2-friendly Data Structure
dds <- DESeqDataSetFromMatrix(countData=counts, colData=meta_data, design=~disease)
#colData is a data frame of demographic/phenotypic data

# filter for genes with significant expression  
# expression should be > 0, but > 100 should be better
dds<-dds[rowSums(counts(dds))>1,] #Gene Filtering

# fit negative binomial regression
dds<-DESeq(dds) #Performs estimation of size factors, dispersion, and negative binomial GLM f#itting

# extracting results, ordering them by adjusted p-value
res <- results(dds)[order(results(dds)[,6]),]
#res[1:10,]

# display results for the top 1000 most significant genes
datatable(data.frame(res[1:1000,]))

# store the name of the top 250 most significant genes
blood.deseq <- rownames(res)[1:250]
blood.res <- data.frame(res)
```

```{r include=FALSE}
# Make a Heatmap of DEGs

# All together: extract the log cpm of the top 250 genes and store that in a matrix
# # get the name of the genes ordered by adjusted p-val
top_genes <- rownames(results(dds)[order(results(dds)$padj),])[1:250]
# # convert log cpm to a matrix structure sorted by the top genes
mat <- as.matrix(assay(se.blood,"log_counts_cpm"))[top_genes,] 

# normalize data row-wise
mat <- t(scale(t(mat)))

# store IDC into a single-column dataframe
df <- data.frame(disease=colData(se.blood)$disease) 

# create HeatMap with some annotations
ha_colors <- list(disease=c("control"="Blue","case"="Red"))

ha <- HeatmapAnnotation(df = df, col = ha_colors)
HM1 <- Heatmap(mat,
               name = "Log CPM",
               show_row_names = F,
               show_column_names = T,
               row_dend_side = "left",
               top_annotation = ha,
               column_title = "Blood Samples",
               heatmap_legend_param = list(direction = "horizontal", 
                                           title_gp = gpar(fontsize = 8), 
                                           labels_gp = gpar(fontsize = 8)))
```

```{r nasal_vars, include=FALSE}
counts <- assay(se.nasal, "counts")
meta_data <- colData(se.nasal)
disease <- meta_data$status
group <- meta_data$status
x <- meta_data$sample_id
labels <- sapply(strsplit(x, "-"), function(parts) {
  paste0(as.numeric(parts[2]), "-", as.numeric(parts[3]))
})
```

```{r include=FALSE}
# fit UMAP model to log cpm data
# note: transpose data to make it tidy
set.seed(1)
umap.defaults$n_neighbors=5
umap_out <- umap(t(assay(se.nasal,"log_counts_cpm")), config=umap.defaults)

# define dataframe with UMAP dimensions and IDC
umap_plot <- as.data.frame(umap_out$layout) # X, Y, ...
umap_plot$disease <- as.factor(se.nasal$disease) # color / group

# plot UMAP1 and UMAP2 by IDC
G2 <- umap_plot %>% ggplot(aes(x=V1, y=V2, color=disease)) +
  geom_point(size=1.5) + xlab("UMAP1") + ylab("UMAP2") +
  geom_text(label=labels, nudge_y = 0.1,) +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("UMAP Plot - Nasal Treatment")
```

## Data Visualization
```{r UMAPS}
combined_plot <- G1 + G2 + plot_layout(guides = "collect") & theme(legend.position = "bottom")

plot(combined_plot)
```

```{r message=FALSE, include=FALSE, cache=TRUE}
# format data into a DESeq2-friendly Data Structure
dds <- DESeqDataSetFromMatrix(countData=counts, colData=meta_data, design=~disease)
#colData is a data frame of demographic/phenotypic data

# filter for genes with significant expression  
# expression should be > 0, but > 100 should be better
dds<-dds[rowSums(counts(dds))>1,] #Gene Filtering

# fit negative binomial regression
dds<-DESeq(dds) #Performs estimation of size factors, dispersion, and negative binomial GLM f#itting

# extracting results, ordering them by adjusted p-value
res <- results(dds)[order(results(dds)[,6]),]
#res[1:10,]

# display results for the top 1000 most significant genes
datatable(data.frame(res[1:1000,]))

# store the name of the top 250 most significant genes
nasal.deseq <- rownames(res)[1:250]
nasal.res <- data.frame(res)
```

```{r}
# Make a Heatmap of DEGs

# All together: extract the log cpm of the top 250 genes and store that in a matrix
# # get the name of the genes ordered by adjusted p-val
top_genes = rownames(results(dds)[order(results(dds)$padj),])[1:250]
# # convert log cpm to a matrix structure sorted by the top genes
mat = as.matrix(assay(se.nasal,"log_counts_cpm"))[top_genes,] 

# normalize data row-wise
mat = t(scale(t(mat)))

# store IDC into a single-column dataframe
df=data.frame(disease=colData(se.nasal)$disease) 

ha = HeatmapAnnotation(df = df, col = ha_colors)
HM2 = Heatmap(mat,
               name = "Log CPM",
               show_row_names = F,
               show_column_names = T,
               row_dend_side = "right",
               top_annotation=ha,
               column_title = "Nasal Samples",
               heatmap_legend_param = list(direction = "horizontal", 
                                           title_gp = gpar(fontsize = 8), 
                                           labels_gp = gpar(fontsize = 8)))
```

## Combined DEGs (DeSeq)
```{r deseq_Heatmaps}
h1 <- grid.grabExpr(draw(HM1,
                         heatmap_legend_side = "bottom",
                         annotation_legend_side = "bottom"))

h2 <- grid.grabExpr(draw(HM2,
                         heatmap_legend_side = "bottom",
                         annotation_legend_side = "bottom"))

# Arrange them separately
grid.arrange(h1, h2, ncol = 2)
```

## Blood Genes
```{r}
#top 1000 list used, pared to 250

ctrl_results <- data.frame(gene = blood.deseq) %>%
    mutate(
        ctrl_padj = blood.res[blood.deseq,]$padj,
        ctrl_effect = blood.res[blood.deseq,]$log2FoldChange
    ) %>% 
    mutate(
        ctrl_effect = ifelse(ctrl_effect > 0, "Upregulated", "Downregulated")
    ) %>% 
    mutate(across(where(is.character), as.factor)) %>% 
    arrange(ctrl_padj) %>% 
    slice_head(n = 250)

row.names(ctrl_results) <- ctrl_results$gene
ctrl_results$gene <- NULL

datatable(ctrl_results)
```

## Blood Pathways

```{r}
showPathways(blood.deseq)
```


## Nasal Genes

```{r}
#top 1000 list used, pared to 250

ctrl_results <- data.frame(gene = nasal.deseq) %>%
    mutate(
        ctrl_padj = nasal.res[nasal.deseq,]$padj,
        ctrl_effect = nasal.res[nasal.deseq,]$log2FoldChange
    ) %>% 
    mutate(
        ctrl_effect = ifelse(ctrl_effect > 0, "Upregulated", "Downregulated")
    ) %>% 
    mutate(across(where(is.character), as.factor)) %>% 
    arrange(ctrl_padj) %>% 
    slice_head(n = 250)

row.names(ctrl_results) <- ctrl_results$gene
ctrl_results$gene <- NULL

datatable(ctrl_results)
```

## Nasal Patheays
```{r}
showPathways(nasal.deseq)
```


## Tuberculosis: Blood VS Nasal - Combined DEGs (DeSeq)
```{r}
common_genes <- intersect(blood.deseq, nasal.deseq)
significance_threshold <- 0.3

combined_results <- data.frame(gene = common_genes) %>%
    mutate(
        blood.padj = blood.res[common_genes,]$padj,
        nasal.padj = nasal.res[common_genes,]$padj,
        combined_chisq = -2*(log(blood.padj) + log(nasal.padj)),
        blood.effect = blood.res[common_genes,]$log2FoldChange,
        nasal.effect = nasal.res[common_genes,]$log2FoldChange,
    ) %>% 
    transmute(
        gene = gene, blood.padj = blood.padj, nasal.padj = nasal.padj,
        combined_p = 1 - pchisq(combined_chisq, 4),
        blood.effect = ifelse(blood.effect > 0, "Upregulated", "Downregulated"),
        nasal.effect = ifelse(nasal.effect > 0, "Upregulated", "Downregulated"),
        reg_agreement = ifelse(blood.effect == nasal.effect, "Same", "Opposite"),
    ) %>% 
    mutate(across(where(is.character), as.factor))%>% 
    filter(
        !is.na(combined_p),
        blood.padj < significance_threshold,
        nasal.padj < significance_threshold,
    ) %>%
    arrange(combined_p) %>% 
    slice_head(n = 250)

row.names(combined_results) <- combined_results$gene
combined_results$gene <- NULL
combined_results$nasal_padj <- NULL
combined_results$ctrl_padj <- NULL


datatable(combined_results)
```

## Nasal VS LPS Stimulated - Combined Pathways

```{r}
showPathways(row.names(combined_results))
```

# Discussion

## Insights

## What next?