---
title: "Analysis of Immune Deficiency RNA-Seq Data"
author: W. Evan Johnson, Ph.D.
date: "`r Sys.Date()`"
header-includes:
   - \usepackage{amsmath}
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    theme: "flatly"
editor_options: 
  chunk_output_type: console
tables: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# load required libraries
library(Rsubread)
library(Rsamtools)
library(SummarizedExperiment)
library(edgeR)
library(DESeq2)
library(ComplexHeatmap)
library(TBSignatureProfiler)
library(umap)
library(tidyverse)
library(DT)
library(GSVA)
library(enrichR)

# define pathway libraries
dbs <- c("WikiPathways_2016","KEGG_2016","Panther_2016","Reactome_2016")
```


### Read in dataset
```{r}

# read RNA expression counts
counts <- read.table("features_combined.txt", sep="\t", 
                     header=T, row.names=1)
# read patient data
meta_data <- read.table("meta_data.txt", sep="\t", 
                     header=T, row.names=1)

# isolate group indicator variable
group <- meta_data$Condition

# aggregate patient and expression data into a single object
se_immunedef <- SummarizedExperiment(assays=list(counts=counts),
                     colData = meta_data)

# create counts per million, log counts and log counts per million features 
# and save them into new assays.
se_immunedef <- mkAssay(se_immunedef, log = TRUE, counts_to_CPM = TRUE)

# display all assays
assays(se_immunedef)
```

### Visualization and Dimension reduction {.tabset}

#### PCA

```{r}
# fit PCA model to log cpm data
# note: transpose data to make it tidy
set.seed(1)
pca_out <- prcomp(t(assay(se_immunedef,"log_counts_cpm")))

# define dataframe with PCA components and immune deficiency condition (IDC)
pca_plot <- as.data.frame(pca_out$x) # X, Y, ...
pca_plot$Condition <- as.factor(se_immunedef$Condition) # color/group

# plot PC1 and PC2 by IDC
g <- pca_plot %>% ggplot(aes(x=PC1, y=PC2, color=Condition)) +
  geom_point(size=1.5) + xlab("PCA1") + ylab("PCA2") +
  geom_text(label=colnames(se_immunedef), nudge_y = 5,) +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("PCA Plot")

plot(g)
```

#### UMAP

```{r}
# fit UMAP model to log cpm data
# note: transpose data to make it tidy
set.seed(1)
umap.defaults$n_neighbors=5
umap_out <- umap(t(assay(se_immunedef,"log_counts_cpm")), config = umap.defaults)

# define dataframe with UMAP dimensions and IDC
umap_plot <- as.data.frame(umap_out$layout) # X, Y, ...
umap_plot$Condition <- as.factor(se_immunedef$Condition) # color / group

# plot UMAP1 and UMAP2 by IDC
g <- umap_plot %>% ggplot(aes(x=V1, y=V2, color=Condition)) +
  geom_point(size=1.5) + xlab("UMAP1") + ylab("UMAP2") +
  geom_text(label=colnames(se_immunedef), nudge_y = 0.1,) +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("UMAP Plot")

plot(g)
```



### Differential Expression {.tabset}

#### DESeq2 Analysis
```{r}
# format data into a DESeq2-friendly Data Structure
dds <- DESeqDataSetFromMatrix(countData = counts, colData=meta_data, design=~Condition)
#colData is a data frame of demographic/phenotypic data

# filter for genes with significant expression  
# expression should be > 0, but > 100 should be better
dds<-dds[rowSums(counts(dds))>1,] #Gene Filtering

# fit negative binomial regression
dds<-DESeq(dds) #Performs estimation of size factors,dispersion, and negative binomial GLM f#itting

# extracting results, ordering them by adjusted p-value
res <- results(dds)[order(results(dds)[,6]),]
#res[1:10,]

# dislpay results for the top 1000 most significant genes
datatable(data.frame(res[1:1000,]))

# store the name of the top 250 most significant genes
deseq250 <- rownames(res)[1:250]
``` 

#### Heatmap of DEGs (DESeq2)

```{r}
# Make a Heatmap of DEGs

# rownames(results(dds)[order(results(dds)$padj),]) # get the name of the genes ordered by adjusted p-val
# as.matrix(assay(se_immunedef,"log_counts_cpm")) # convert log cpm to a matrix structure
# All together: extract the log cpm of the top 250 genes and store that in a matrix
mat = as.matrix(assay(se_immunedef,"log_counts_cpm"))[rownames(results(dds)[order(results(dds)$padj),])[1:250],] # Using first 500 genes to simplify

# normalize data row-wise
mat = t(scale(t(mat)))

# store IDC into a single-column dataframe
df=data.frame(Condition=colData(se_immunedef)$Condition) 

# create HeatMap with some annotations
ha = HeatmapAnnotation(df = df, col = list(Condition=c("Deficient"="Red","Control"="Blue")))
Heatmap(mat,show_row_names=F,show_column_names = F, top_annotation = ha)

```


#### Limma Analysis

```{r}
# Store data in Limma-friendly Data Structure
dge <- DGEList(counts=counts, group=group) #From edgeR, Computes library size

# filter for genes with significant expression
counts<-counts[which(rowSums(cpm(counts))>1),] #Gene Filtering #(mask)
dge <- DGEList(counts=counts, group=group) #Re-compute library size #(filter)

# Trimmed-Mean Normalization of data
dge <- calcNormFactors(dge) #TMM normalization

# Create design matrix (intercept + indicator variable)
design<-model.matrix(~group)

# distribution normalizing transformation
v<-voom(dge, design) #voom transform to calculate weights to eliminate mean-variance #relationship
#use usual limma pipelines

# fit empirical bayes regression model
fit<-lmFit(v,design)
fit<-eBayes(fit)

# organize data by top ranked genes when predicting IDC (coef = col#2 = IDC)
topTable(fit, coef=ncol(design))

# display top 1000 most significant genes
datatable(topTable(fit, coef=ncol(design), number=1000))

# store the name of the top 250 most significant genes
limma250 <- rownames(topTable(fit, coef=ncol(design), number=250))
```

#### Heatmap of DEGs (limma)

```{r}
# Make a Heatmap of DEGs

# rownames(topTable(fit, coef=ncol(design), number=50)) # get the name of the top 50 most significant genes
# as.matrix(assay(se_immunedef,"log_counts_cpm")) # convert log cpm to a matrix structure
# All together: extract the log cpm of the top 50 genes and store that in a matrix
mat = as.matrix(assay(se_immunedef,"log_counts_cpm"))[rownames(topTable(fit, coef=ncol(design), number=50)),] # Using first 500 genes to simplify

# normalize data row-wise
mat = t(scale(t(mat)))

# store IDC into a single-column dataframe
df=data.frame(Condition=colData(se_immunedef)$Condition) 

# create HeatMap with some annotations
ha = HeatmapAnnotation(df = df, col = list(Condition=c("Deficient"="Red","Control"="Blue")))
Heatmap(mat,show_row_names=T,show_column_names = F, top_annotation = ha)
```




### Pathway analysis {.tabset}

#### Fetch C7 immune genesets

```{r, eval=F}
library(msigdbr)

# retrieve immune-related gene sets from the Molecular Signatures Database
immune_gene_sets = msigdbr(species = "human", category = "C7", subcategory = "IMMUNESIGDB")

# extract the pathway names (no dupliactes)
path_names <- unique(immune_gene_sets$gs_name)
length(path_names) # pathway count

# function to get all the gene names belonging to a specific pathway
get_genes<- function(pname, sets){sets %>% filter(gs_name==pname) %>% pull(gene_symbol)}

# create a list of pathways as keys and vector of belonging genes as values
C7_pathlist <- sapply(path_names, get_genes, sets)

# save list
saveRDS(C7_pathlist, file="mSigDB_C7.rds")
```


#### Score C7 Immune pathways
```{r}
# load list containing immune-related pathways as keys
# and vectors of the immune-related genes belonging to that pathway
C7_pathlist <- readRDS(file="mSigDB_C7.rds")

library(GSVA)

# package data into GSVA-friendly format:
#  - genes x patient log cpm expression measures
#  - gene pathway data
gsvaPar <- gsvaParam(assay(se_immunedef,"log_counts_cpm"), C7_pathlist)

# run GSVA (Gene Set Variation Analysis) 
# computing enrichment scores per sample for each immune-related pathway
# result: matrix of pathway x patient enrichment scores
gsvaRes <- gsva(gsvaPar, verbose=FALSE)
```

#### Limma Analysis

```{r}
# Create design matrix (intercept + indicator variable)
design<-model.matrix(~group)

# fit empirical bayes regression model
#relationship
#use usual limma pipelines
fit<-lmFit(gsvaRes,design)
fit<-eBayes(fit)

# organize data by top ranked pathways when predicting IDC (coef = col#2 = IDC)
topTable(fit, coef=ncol(design))

# display top 100 ranked pathways when predicting IDC
datatable(topTable(fit, coef=ncol(design), number=1000))
```

#### Heatmap of DEGs (limma)

```{r}
# rownames(topTable(fit, coef=ncol(design), number=250)) # get the name of the top 250 most significant pathways
# as.matrix(gsvaRes) # convert enrichment scores to a matrix structure
# All together: extract the enrichment scores of the top 250 pathways and store that in a matrix

# Make a Heatmap of DEGs
mat = as.matrix(gsvaRes)[rownames(topTable(fit, coef=ncol(design), number=250)),] # Using first 500 genes to simplify

# normalize data row-wise
mat = t(scale(t(mat)))

# store IDC into a single-column dataframe
df=data.frame(Condition=colData(se_immunedef)$Condition) 

# create HeatMap with some annotations
ha = HeatmapAnnotation(df = df, col = list(Condition=c("Deficient"="Red","Control"="Blue")))
Heatmap(mat,show_row_names=F,show_column_names = F, top_annotation = ha)

```
 
#### EnrichR analysis {.tabset}

##### DESeq2 Genes {.tabset}
 
```{r}
# based on the results of DESeq, cross-reference the gene with pathway databases to infer which pathways are the most significant
a <- enrichr(deseq250, dbs)

# next code-chunks: display the results of each database cross-referencing
```


###### DESeq2 WikiPathways
```{r}
datatable(enriched$WikiPathways_2016)
```

###### DESeq2 KEGG
```{r}
datatable(enriched$KEGG_2016)
```

###### DESeq2 Panther
```{r}
datatable(enriched$Panther_2016)
```

###### DESeq2 Reactome
```{r}
datatable(enriched$Reactome_2016)
```

##### Limma Genes {.tabset}
 
```{r}
# based on the results of limma, cross-reference the gene with pathway databases to infer which pathways are the most significant
enriched <- enrichr(limma250, dbs)

# next code-chunks: display the results of each database cross-referencing
```

###### Limma WikiPathways
```{r}
datatable(enriched$WikiPathways_2016)
```

###### Limma KEGG
```{r}
datatable(enriched$KEGG_2016)
```

###### Limma Panther
```{r}
datatable(enriched$Panther_2016)
```

###### Limma Reactome
```{r}
datatable(enriched$Reactome_2016)
```